---
title: "Volcano Plot using ggplot2"
author: "Debojyoti Das"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

## Introduction

This document demonstrates how to generate a volcano plot using `ggplot2` by reading a CSV file that contains gene expression data. The dataset must include at least three mandatory columns:

- `log2FC` (Log2 Fold Change)
- `p_value` (P-value for statistical significance)
- `Gene_symbol` (or Gene EntrezID or Gene ENSEMBL ID)

Each step is explained in detail, with code chunks for clarity.

## Installing and Loading Required Libraries

Check if required packages are installed, if not install them
```{r install_libraries, echo=TRUE}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}

if (!requireNamespace("ggrepel", quietly = TRUE)) {
  install.packages("ggrepel")
}

if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}
```

Load libraries
```{r load_libraries, echo=TRUE}
library(ggplot2)
library(dplyr)
library(ggrepel)
```

## Setting Up Project Directories

```{r directory_setup, echo=TRUE}
project_dir <- "/Users/debda22/Projects/core_facility_projects/ggplot_basics"
data_dir <- paste0(project_dir, "/input_data")
result_dir <- paste0(project_dir, "/results")
```

Create directories if they don't exist
```{r setup, echo=TRUE}
if (!dir.exists(result_dir)) dir.create(result_dir, recursive = TRUE)
```

## Reading the Input Data

```{r read_data, echo=TRUE}
input_file <- paste0(data_dir, "/test_input_file.csv")
data <- read.csv(input_file)
```

Display first few rows of the dataset
```{r check_data, echo=TRUE}
head(data)
```

## Transforming Data for Visualization

Before plotting, we transform the data:
- Convert the `p_value` column to `-log10(p_value)` to emphasize small p-values.
- Define upregulated and downregulated genes based on cutoff values.
- Reverse the `log2FC` values for visualization.

We define thresholds for classification:
- `p_value` cutoff: 0.05
- `log2FC` cutoff: 1


Define cutoffs
```{r set_cutoff, echo=TRUE}
pval_cutoff <- 0.05
log2fc_cutoff <- 1
```

Classify genes into upregulated, downregulated, or Non-significant
```{r classify_data, echo=TRUE}
data <- data %>%
  mutate(
    logP = -log10(p_value),
    negLog2FC = -log2FC,
    regulation = case_when(
      p_value < pval_cutoff & log2FC > log2fc_cutoff ~ "Upregulated",
      p_value < pval_cutoff & log2FC < -log2fc_cutoff ~ "Downregulated",
      TRUE ~ "Non-significant"
    )
  )
```

## Selecting Top Genes for Labeling

To highlight important genes, we select the top `n` genes from the upregulated and downregulated groups.

```{r select_top_genes, echo=TRUE}
top_n <- 5  # Number of genes to label

top_up <- data %>%
  filter(regulation == "Upregulated") %>%
  arrange(desc(log2FC)) %>%
  head(top_n)

top_down <- data %>%
  filter(regulation == "Downregulated") %>%
  arrange(desc(abs(log2FC))) %>%
  head(top_n)

# Combine top genes
top_genes <- bind_rows(top_up, top_down)
```

## Creating the Volcano Plot

We use `ggplot2` to create a volcano plot with color distinctions:

```{r generate_volcano, echo=TRUE}
volcano <- ggplot(data, aes(x = negLog2FC, y = logP, color = regulation)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("Non-significant" = "black", "Upregulated" = "red", "Downregulated" = "blue")) +
  labs(title = "Volcano Plot", x = "-Log2 Fold Change", y = "-Log10 P-value") +
  theme_minimal() +
  geom_text_repel(
    data = top_genes, 
    aes(label = Gene_symbol), 
    vjust = -1, 
    size = 3, 
    show.legend = FALSE
    ) +
  ylim(c(0,6))
```

Display plot
```{r plot_volcano, echo=TRUE}
print(volcano)
```

## Saving the Plot

```{r save_plot, echo=TRUE}
output_file <- paste0(result_dir, "/volcano_plot.png")
ggsave(output_file, plot = volcano, width = 8, height = 6)
```

## Conclusion

This document demonstrated how to load, process, and visualize gene expression data using a volcano plot. We added classification for upregulated and downregulated genes, highlighted top genes, and saved the final plot.
